The results support three main conclusions.

\paragraph{Exact second-order information is worthwhile, but not everywhere in
the same way.}
In the scalar benchmark families reported here, exact element Hessians were the
cleanest and often fastest choice. On \plasticitythree{}, exact constitutive
tangents were the practical winner because they preserve the energy-based AD
story while avoiding the cost of full element Hessians on high-order
tetrahedra. On other problems, colored SFD remains methodologically useful
because it gives a structured approximation without requiring a hand-derived
tangent.

\paragraph{The best preconditioner is not always the one with the fewest Krylov
iterations.}
Several comparison studies in the repository illustrate this point, and the
paper keeps the conclusion explicit. Some PMG profiles reduce iteration counts
while still losing on wall-clock time because each apply is much more
expensive. In the \plasticitythree{} comparisons reported here, lower Krylov
counts did not always translate into lower wall-clock time. This is why the
paper emphasizes both iteration counts and component timings.

\paragraph{Reference implementations remain essential.}
The \fenics{} and \purejax{} paths are not just historical leftovers. They make
the mainline \jaxpetsc{} story more interpretable by exposing where the
maintained production stack is genuinely adding something new and where it is
merely reproducing a smaller reference implementation at larger scale. The
benchmark families also make clear that references are unevenly available:
topology has a strong \purejax{} serial reference but no maintained \fenics{}
counterpart; \plasticitythree{} has source-assembly engineering comparisons but
not a maintained \fenics{} or \purejax{} baseline.

There are also limitations that the paper should state plainly. The topology
fine-grid scaling study is end-to-end, not fixed-work. The 2D plasticity large
scaling campaign is intentionally capped at a fixed Newton budget. Some of the
source/local PMG profile comparisons remain engineering diagnostics rather than
fully symmetric scientific baselines. None of these caveats invalidate the
results, but they do affect how claims should be phrased.

The broader methodological lesson is that differentiable FEM is most useful
when it is paired with solver-aware engineering. Automatic differentiation by
itself does not solve nonlinear FEM. Sparse layout, preconditioning, Krylov
policy, globalization, and distributed ownership rules are all part of the
final scientific behavior. The main contribution of \repo{} is to show that
these pieces can be kept together in one maintained implementation surface.
