Many finite-element software stacks provide a strong subset of the ingredients
needed for nonlinear solves and optimization, but relatively few combine all of the
following in a maintained and practically scalable way: expressive high-level
problem definitions, exact or near-exact gradients, usable second-order
information, distributed nonlinear solves, and preconditioning strategies that
remain viable on hard mechanics and design problems. This gap becomes most
visible in the benchmark families that motivated \repo{}: high-order
Mohr--Coulomb plasticity in three dimensions, nonlinear hyperelasticity, and
topology optimization with coupled design and mechanics updates.

The repository therefore treats \jaxpetsc{} as the main production path while
keeping \fenics{} and \purejax{} baselines wherever they are scientifically
useful. The \fenics{} implementations act as trusted PDE references on scalar
and elasticity families. The \purejax{} paths provide serial differentiable
references. The maintained \jaxpetsc{} stack is where we aim to integrate the
whole story: sparse distributed vectors and matrices, nonlinear globalization,
multiple derivative modes, and preconditioned Krylov solvers.

The central claim of the paper is intentionally practical. We do not claim that
every benchmark requires exact Hessians, nor that every mainstream framework is
unable to provide them. Instead, we show that exact gradients and second-order
information can be made operational inside a distributed nonlinear FEM stack,
and that this materially changes what can be benchmarked and compared. In the
maintained repository this leads to three concrete capabilities:
\begin{enumerate}
  \item exact element-level derivatives when they are affordable;
  \item constitutive-level automatic differentiation for high-order plasticity
        when full element Hessians are too expensive;
  \item colored sparse finite-difference (SFD) recovery as a scalable fallback
        when exact second-order information is not the best tradeoff.
\end{enumerate}

These capabilities are evaluated on six maintained families:
$p$-Laplace, Ginzburg--Landau, hyperelasticity, plasticity, \plasticitythree{},
and topology optimization. We give them equal top-level space because the
paper is about a software methodology rather than one bespoke solver. At the
same time, we devote extra narrative attention to \plasticitythree{} and
topology optimization. Those two families are where the implementation
difficulty is highest and where the case for the maintained \jaxpetsc{}
approach is strongest. For \plasticitythree{}, the combination of exact
constitutive tangents, multigrid, and MPI-parallel Newton solves is the
principal result. For topology optimization, the challenge is different: a
distributed design update must coexist with a robust mechanics solve and a
continuation schedule that is numerically useful without becoming impossible to
reproduce.

The paper is organized as follows. \Cref{sec:related} places the work in the
context of differentiable FEM, adjoint-based PDE optimization, and PETSc
nonlinear solvers. \Cref{sec:methodology} describes the mathematical and
algorithmic ingredients, including line search, trust regions, hybrid
globalization, and colored derivative recovery. \Cref{sec:implementation}
describes how these ideas are mapped onto \purejax{}, \fenics{}, and
\jaxpetsc{}. \Cref{sec:benchmarks} introduces the benchmark suite, while
\Cref{sec:numerical-examples} presents the later timing, scaling, and
discretisation-comparison evidence. We conclude with a discussion of where
exact second-order information helps most and where cheaper approximations
remain the better engineering choice.

The manuscript is written in an arXiv-first style and uses the current repo
artifacts as the canonical data source. This is deliberate. Our goal is to make
the scientific narrative and the implementation surface align closely enough
that the paper can act as a map for both readers and future maintainers.
