\label{sec:related}

The paper sits at the intersection of automated finite-element workflows,
PDE-constrained optimization, differentiable simulation, and scalable nonlinear
linear algebra. The most relevant background can be grouped into four strands.

\paragraph{High-level finite-element automation.}
The \fenics{} project established a powerful model in which weak forms are
expressed at a high level and compiled into efficient low-level kernels
\citep{logg2012fenicsbook}. That abstraction has been enormously influential.
For our purposes it remains important both as a reference implementation path
and as a baseline for what an expressive PDE programming environment should
feel like.

\paragraph{Adjoint-based PDE optimization.}
In the \fenics{}/Firedrake ecosystem, \code{dolfin-adjoint} and
\code{pyadjoint} showed that many PDE-constrained optimization problems can be
treated through automated adjoints at the level of high-level problem
descriptions \citep{farrell2013dolfinadjoint,mitusch2019pyadjoint}. These tools
are especially strong for first-order reduced sensitivities. They are an
important scientific reference point for this paper because they demonstrate
how much can be automated without exposing the user directly to low-level
linear algebra. At the same time, they also illustrate our main distinction:
the combined package of exact second-order information, distributed nonlinear
solves, and difficult material models is not commonly surfaced in the same
maintained workflow.

\paragraph{Differentiable simulation and JAX-native FEM.}
The use of automatic differentiation inside simulation codes is now widespread,
and the broader AD literature is mature
\citep{baydin2018autodiff,nocedal2006numerical}. JAX, in particular, has made
higher-order derivatives, vectorization, and accelerator-friendly program
transformations accessible in a NumPy-like environment \citep{jax2018}. This
has encouraged differentiable FEM frameworks such as JAX-FEM, which show that
nonlinear constitutive models and inverse-design workflows can be expressed
effectively in JAX \citep{xue2023jaxfem}. Our work is closest in spirit to this
strand, but differs in emphasis: the mainline implementation here is not just a
differentiable solver, but a distributed Newton--Krylov + PETSc workflow whose
data structures are built around sparse MPI execution.

\paragraph{Topology optimization and nonlinear mechanics.}
The topology optimization literature is rich, with classical introductions and
compact educational implementations still serving as community reference points
\citep{sigmund2001topology,bendsoe2003topology,ferrari2020top99}. For
inelasticity and nonlinear mechanics, consistent tangents and robust
globalization remain standard themes \citep{simo1985consistent,simo1998compinel}.
Our 3D plasticity implementation is not a textbook return-mapping framework,
but these references are still important because they clarify what a nonlinear
mechanics code must preserve to make Newton-type methods meaningful.

\paragraph{Sparse derivative estimation and coloring.}
When exact Hessians become too expensive, sparse recovery via graph coloring is
the classical alternative. Curtis, Powell, and Reid established the key idea
for sparse Jacobians \citep{curtis1974sparsejacobian}; Coleman and Mor\'e
extended the viewpoint to sparse derivative estimation and graph-coloring-based
recovery \citep{coleman1983sparsejacobian,coleman1984sparsehessian}. These
ideas remain directly relevant here because the maintained \jaxpetsc{} stack
uses local and distributed coloring for SFD Hessian assembly.

\paragraph{Scalable nonlinear and linear algebra.}
PETSc provides the scalable vector, matrix, Krylov, nonlinear, and multigrid
infrastructure that makes our distributed implementation possible
\citep{balay1997petsc,petsc2024web}. Rather than treating PETSc as a back-end
detail, we treat it as a first-class methodological component: solver policy,
preconditioner choice, and distributed data layout are part of the scientific
story, not just engineering glue.

\begin{table}[H]
  \centering
  \input{tables/generated/sota_framework_comparison.tex}
  \caption{Positioning of the manuscript relative to nearby framework families
  and workflow styles. ``Partial'' denotes that some form of the capability is
  available in the cited ecosystem, but not necessarily as an exact
  second-order, distributed nonlinear workflow for the same benchmark class.}
  \label{tab:sota-comparison}
\end{table}

Against this background, the niche of \repo{} can be stated precisely. It is a
methods-and-software framework that tries to occupy the middle ground between
high-level differentiable modeling and scalable sparse nonlinear solvers. The
paper does not claim that this is the only worthwhile point in that design
space. It argues that this point is underrepresented, practically useful, and
scientifically worth documenting in detail.
